---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Let's load the data by using "read.csv" function.
```{r}
unzip("activity.zip")
df = read.csv("activity.csv")
```

So, we have loaded the dataframe, let's check out the columns.
```{r}
str(df)
```

As we can see the date column is in character class, let's change its datatype into date.
```{r}
df[,2] = as.Date(df[,2])
str(df)
```

## What is mean total number of steps taken per day?
Now let's calculate the total number of steps taken per day.
```{r}
dftotal = aggregate(df[,1],list(df[,2]), FUN = sum)
```

After getting the total steps taken per day, we will now plot the histogram of the new dataframe.
```{r}
Freq <- dftotal[,2]
for(i in seq_along(Freq)){
    if(is.na(Freq[i])){
    Freq[i]=0 #replacing NA values with 0
    }
}
Overall.Cond <- 1:(length(Freq)+1)
myhist <-list(breaks=Overall.Cond, counts=Freq, density=Freq/diff(Overall.Cond),xname="Days(from 1st Oct to 30th Nov)", ylab="No of steps")
class(myhist) <- "histogram"
plot(myhist, main = "Histogram of daily steps-1")
```

Now we have to calculate and report the mean and median of the total number of steps taken per day.
```{r}
dfc = df
for(i in seq_along(dfc[,1])){
    if(is.na(dfc[i,1])){
        dfc[i,1]=mean(df[,1],na.rm = TRUE)
    }
}
dfavg = aggregate(dfc[,1],list(dfc[,2]), FUN = mean)
dfmedian = aggregate(dfc[,1],list(dfc[,2]), FUN = median)
dfavg = cbind(dfavg,dfmedian[,2])
names(dfavg) = c("Date","Mean","Median")
dfavg
```

## What is the average daily activity pattern?
This time, we will find the steps per each 5 minute interval averaged across all days.
```{r}
df5min = aggregate(dfc[,1],list(dfc[,3]), FUN = mean)
names(df5min)[1] = 'Interval'
head(df5min)
```

So as we have git the average steps per each interval dataset, let's plot it.
```{r}
plot(df5min[,1],df5min[,2], type='l',xlab="Interval",ylab="Average Steps", main = "Plot of steps per each interval")
```

Now let's find which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.
```{r}
ansmin = df5min[(df5min[,2]==max(df5min[,2])),1]
ansmin
```

Thus we found that at interval ``r ansmin`` there is maximum number of steps.

## Imputing missing values
Now let's count the total number of missing values in the dataset.
```{r}
a=sum(is.na(df[,1]))
b=sum(is.na(df[,2]))
c=sum(is.na(df[,3]))
c(a,b,c)
```

So there are ``r a`` rows in the dataframe with NAs.
To clean the dataset, let's replace the NAs with the median value of steps for corresponding day.
```{r}
df2 = df
for(i in seq_along(df2[,1])){
    if(is.na(df2[i,1])){
        subset2 = subset(df2,interval==df[i,3])
        df2[i,1]=as.integer(mean(subset2[,1],na.rm = TRUE))
    }
}
dfavg2 = aggregate(df2[,1],list(df2[,2]), FUN = mean)
dfmedian2 = aggregate(df2[,1],list(df2[,2]), FUN = median)
dfavg2 = cbind(dfavg2,dfmedian2[,2])
names(dfavg2) = c("Date","Mean","Median")
dfavg2
```
As we have calculated the means and medians of total number of steps taken per day, let's make a histogram of the total number of steps taken each day.
```{r}
dftotal2 = aggregate(df2[,1],list(df2[,2]), FUN = sum)
Freq <- dftotal2[,2]
sum(is.na(Freq))
Overall.Cond <- 1:(length(Freq)+1)
myhist <-list(breaks=Overall.Cond, counts=Freq, density=Freq/diff(Overall.Cond),xname="Days(from 1st Oct to 30th Nov)", ylab="No of steps")
class(myhist) <- "histogram"
plot(myhist, main = "Histogram of daily steps-2")
```

The difference between the plots with the 2 different kinds of imputations is not quite noteworthy. 

## Are there differences in activity patterns between weekdays and weekends


Now let's create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
df2[,4]=weekdays(df2[,2])
dayz = unique(df2[,4])
for(i in seq_along(dayz)){
  if(i<6){
    df2[,4]=gsub(dayz[i],"weekday",df2[,4])
  }
  else{
    df2[,4]=gsub(dayz[i],"weekend",df2[,4])
  }
}
df2$V4 = as.factor(df2$V4)
names(df2)[4] = "daytype"
```

Having created the 4th column of the day-type, let's make 2 plots of  the average number of steps taken across all weekday days and weekend days per all 5-minute interval.

```{r}
library(ggplot2)
dfag = aggregate(df2[,1],list(df2[,3],df2[,4]),FUN=mean)
qplot(Group.1,x,data=dfag,facets = 'Group.2',geom = 'line',xlab = 'Interval',ylab = 'No. of steps')
```

We can notice that on weekend days the average steps are more than weekday days, while during peak hours, the number of steps taken is higher on weekdays.